---
title: "Assignment"
author: "Jussi Kauppinen 2891068547"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

# Task 1

a) The reason why normal distribution is so special in context of general linear models is statistical inference. Usually in case of general linear models the confidence intervals and hypothesis testing for the model parameters are important parts of the process. In cases where random variables error terms follow the normal distribution with same variance the exact distribution for the MLE of parameter vector can be determined. The MLE is actually normally distributed. This makes statistical inference simple. In cases where variances are not constant similar results arise. In cases where error terms are not normally distributed the sampling distributions can be obtained with asymptotic distributions. For large samples the standardized estimator can be approximated by normal distribution.

b) Probability density function of the $\chi^2$-distribution is

\begin{equation}
\begin{split}
f(y, k) &= \frac{1}{2^{k/2} \Gamma(k/2)} y^{k/2 - 1} e^{-y/2}\\
&= \text{exp} \left( \log \left( \frac{1}{2^{k/2} \Gamma(k/2)} y^{k/2 - 1} e^{-y/2} \right) \right)\\
&= \text{exp} \left( - \frac{k}{2} \log 2 - \log \Gamma(\frac{k}{2}) + (\frac{k}{2} - 1) \log y - \frac{y}{2} \right)\\
&= \text{exp} \left( (\frac{k}{2} - 1) \log y - (\frac{k}{2} \log 2 + \log \Gamma(\frac{k}{2})) -\frac{y}{2} \right)\\
\end{split}
\end{equation}

Let's set $\theta = \frac{k}{2}-1$, $b(\theta) = (\theta+1) \log2 + \log \Gamma(\theta+1)$, $c(y, \phi) = -\frac{y}{2}$, $\phi = 1$ and $a(\phi)=1$. We can see that this is almost fulfills the form of exponential family. Only thing is that we have $\log y$ instead of $y$. Apparently this is sufficient and $\chi^2$-distribution is a member of exponential family. Canonical link is $\theta = \frac{k}{2}-1$. However the form of the distribution yields following results:

$$E(\log(y)) = b'(\theta)$$
$$Var(\log(y)) = a(\phi)b''(\theta)$$
I think this might be the reason why glm-function in R doesn't support $\chi^2$-distribution.


# Task 2

Hurdle model consists of two components. One component for positive counts such as truncated Poisson model and one hurdle component for zero versus larger counts. [@hurdle]. The hurdle models probability density function is

\begin{equation}
f_{hurdle}(y;x,z,\beta, \gamma) = 
\begin{cases}
f_{zero}(0;z,\gamma) &, \quad \text{if } y=0\\
(1-f_{zero}(0;x,\gamma)) \frac{f_{count}(y;x,\beta)}{1-f_{count}(0;x,\beta)} &, \quad \text{if } y>0\\
\end{cases}
\end{equation}

From here we can see that the log-likelihood function will be

\begin{equation}
\begin{split}
l(y;x,z,\beta,\gamma) &= \sum_{i=1}^k \log f_{zero}(0;z_i,\gamma) + \sum_{i=k+1}^n \log(1 - f_{zero}(0;x_i,\gamma)) + \log f_{count}(y_i;x_i;\beta) - \log(1 - f_{count}(0;x_i;\beta))\\
\end{split}
\end{equation}

assuming that the first $k$ observations are $0$. So the model parameters can be solved separately from estimating equations:

\begin{equation}
\begin{split}
l'(\gamma) &= \frac{\partial}{\partial \gamma} \left(\sum_{i=1}^k \log f_{zero}(0;z_i,\gamma) + \sum_{i=k+1}^n \log(1 - f_{zero}(0;x_i,\gamma))\right) = 0\\
\end{split}
\end{equation}

and

\begin{equation}
\begin{split}
l'(\beta) &= \frac{\partial}{\partial \beta} \left(\sum_{i=k+1}^n \log f_{count}(y_i;x_i;\beta) - \log(1 - f_{count}(0;x_i;\beta))\right) = 0\\
\end{split}
\end{equation}

```{r message=FALSE, warning=FALSE}
library(glmmTMB)
library(pscl)
library(ggplot2)
library(lattice)
library(mltest)
library(ResourceSelection)
library(xtable)
```

Let's demonstrate different models with Salamanders data from glmmTMB-package [@Salamanders1] [@Salamanders2]. Let's choose some example model to demonstrate the observed number of salamanders. 

```{r}
salam <- Salamanders
attach(salam)

plot(table(count), ylab="frequency")

#Mining's effect on the observed number of salamanders
plot(count ~ mined)

#Covers effect on the observed number of salamanders
plot(count ~ cover)
```

It seems like mountain top removal mining affects the number of salamanders on sites. Also the amount of cover objects also seems to have an effect. It can also be seen that there's lots of zero-observations which calls for hurdle model.

```{r}
#Regular Poisson model
fit <- glm(count ~ mined + cover, data = salam, family="poisson")
summary(fit)

#Hurdle model
fith <- hurdle(count ~ mined + cover, data = salam,
               dist="poisson", zero.dist="binomial", link="logit")
summary(fith)
```

Let's compare maximum log-likelihoods and AIC-values of the models and compare what kind to estimations both models give for number of zero-observations.

```{r}
#AIC for both models
AIC(fit, fith)

#Compare log-likelihoods
logLik(fit)
logLik(fith)

#Expected number of zero-observations from both models and the actual number.

#Poisson
sum(dpois(0, fitted(fit)))

#Hurdle
sum(predict(fith, type="prob")[,1])

#Actual
sum(as.numeric(salam$count==0))
```
We can see from all this criteria that hurdle model is better in this case. For the interpretations here the hurdle component basically gives the probability to encounter any salamanders. The count component models the positive counts of lizards. Let's make some interpretations from the models.

```{r}
b1 <- exp(coef(fit))
b1

b2 <- exp(coef(fith))
b2

#Expected number of observed salamanders when the site was affected by
#mountain top removal coal mining and scaled number of cover objects is
#zero.

#Poisson
b1[1]

#Hurdle
b2[1]

mean(subset(salam, salam$mined=="yes")$count)
```

In this case it seems like Poisson model gives a better interpretation. Hurdle models value 1,25 looks quite high. However I think there might be some thing about the truncation that would modify the result a bit.

```{r}
#Percentual change in observed number of salamanders in sites where other is
#affected by mining and other is not.

#Poisson
(b1[2]-1)*100

#Hurdle
(b2[2]-1)*100
```

Both models give a result that mountaintop removal affects observed number of salamanders negatively but the Poisson model gives a result that seems unrealistically high.

```{r}
#Percentual change in number of observed salamanders between sites with one 
#unit difference in scaled number of cover objects.

#Poisson
(b1[3]-1)*100

#Hurdle
(b2[3]-1)*100
```

Effect of cover objects seems to be similar in both models. The hurdle model can be also interpreted from it's zero vs. positive count part as odds and odds ratios since logit-link was used.

```{r}
#Odds to observe any salamanders in a site that was affected by mining
#and scaled number of cover objects is zero.
b2[4]

#OR between site that is affected by mining and site that isn't.
b2[5]

#OR between sites with one unit difference in scaled number of cover objects.
b2[6]
```

So it is rather likely to not observe any salamanders in a site. It is much more likely to encounter any salamanders in a site that is not affected by mining. It is also more likely to encounter salamanders when there are less cover objects.

# Task 3

```{r}
discharge <- read.table("discharge.txt", header=TRUE, sep=";")

#Shift binary values to 0 and 1
discharge$death <- discharge$death - 1
discharge$gender <- discharge$gender - 1
discharge$inh_inj <- discharge$inh_inj - 1
discharge$flame <- discharge$flame - 1
discharge$race <- discharge$race - 1
discharge$gender <- factor(discharge$gender, levels=c(0,1), 
                           labels=c("Female", "Male"))
discharge$inh_inj <- factor(discharge$inh_inj, levels=c(0,1), 
                            labels=c("No", "Yes"))
discharge$flame <- factor(discharge$flame, levels=c(0,1), 
                          labels=c("No", "Yes"))
discharge$race <- factor(discharge$race, levels=c(0,1), 
                         labels=c("Non-white", "White"))

#Convert age and tbsa to numeric values
discharge$age <- as.numeric(gsub(",", ".", discharge$age))
discharge$tbsa <- as.numeric(gsub(",", ".", discharge$tbsa))

set.seed(8547)

#sample indexes
s <- sample(nrow(discharge), 850)

#Split discharge data into modeling data and test data
data <- discharge[s,]
test <- discharge[-s,]



breaks <- c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90)

labels <- c("0-10", "10-20", "20-30", "30-40", "40-50", "50-60", "60-70", "70-80", "80-90")

#Age-groups for visualization
data$age_group <- cut(data$age, breaks = breaks, labels = labels, right = FALSE)

#logit-function to investigate linearity of the effects
logit <- function(data){
  log(mean(data)/(1-mean(data)))
}

#Compute ratio of deaths vs cases for each age-group.
dratioage <- aggregate(data$death, by=list(data$age_group), FUN=logit)

ggplot(data=dratioage, aes(x=Group.1, y=x))+
  geom_point()+
  xlab("Age")+
  ylab("log-odds of dying")+
  theme_bw()
```

Relationship of age to odds of dying seems to be relatively linear.

```{r}
breaks <- seq(0,100, 10)

labels <- c("0-10", "10-20", "20-30", "30-40",
            "40-50", "50-60", "60-70", "70-80",
            "80-90", "90-100")

data$tbsa_group <- cut(data$tbsa, breaks = breaks, labels=labels, right = FALSE)

dratioburn <- aggregate(data$death, by=list(data$tbsa_group), FUN=logit)

ggplot(data=dratioburn, aes(x=Group.1, y=x))+
  geom_point()+
  xlab("Total burn area")+
  ylab("log-odds of dying")+
  theme_bw()
```

Total surface burn area has obvious effect on probability of dying. Relationship looks somewhat linear.

```{r}
dratioinh <- aggregate(data$death, by=list(data$inh_inj), FUN=mean)

ggplot(data=dratioinh, aes(x=as.factor(Group.1), y=x))+
  geom_bar(stat="identity")+
  xlab("Burn involved inhalation injury")+
  ylab("Mortality rate")+
  ylim(0,1)
```

Inhalation injury has major effect on mortality rate.

```{r}
dratioflame <- aggregate(data$death, by=list(data$flame), FUN=mean)

ggplot(data=dratioflame, aes(x=as.factor(Group.1), y=x))+
  geom_bar(stat="identity")+
  xlab("Flame involved in burn injury")+
  ylab("Mortality rate")+
  ylim(0,1)
```

Flame being involved in injury has some effect too. However both flame involved in injury and inhalation injury can be both correlated with the total burnt surface area.

```{r}
ggplot(data=data, aes(x=inh_inj, y=tbsa))+
  geom_boxplot()

ggplot(data=data, aes(x=flame, y=tbsa))+
  geom_boxplot()
```

It seems like patients with inhalation injuries also have much more total burnt surface area which makes sense. Inhalation injury seemed like it has major effect on mortality on it's own so it might be worth to keep it in the final model. Flame being involved didn't have that much effect on mortality and it doesn't seem that of a relevant variable here.

```{r}
dratiogender <- aggregate(data$death, by=list(data$gender), FUN=mean)

ggplot(data=dratiogender, aes(x=as.factor(Group.1), y=x))+
  geom_bar(stat="identity")+
  xlab("Gender")+
  ylab("Mortality rate")+
  ylim(0,1)

dratiorace <- aggregate(data$death, by=list(data$race), FUN=mean)

ggplot(data=dratiorace, aes(x=as.factor(Group.1), y=x))+
  geom_bar(stat="identity")+
  xlab("Race")+
  ylab("Mortality rate")+
  ylim(0,1)
```

Gender and race seem to have little to no effect to mortality rate.

Let's begin by fitting a model with all variables.

```{r}
fitall <- glm(death ~ age + gender + race + tbsa + inh_inj + flame, data=data,
              family="binomial")
summary(fitall)
```

Significance of gender and race is in line with what we saw earlier so they are dropped from the model. Flame is also dropped since it seems like it is not significant. For interpretations age is centered to 30 years and tbsa to 15 %. Let's see if interaction with inhalation injury is needed.

```{r}
fit1 <- glm(death ~ I(age-30)*inh_inj + I(tbsa-15)*inh_inj, data=data,
           family="binomial")

fit2 <- glm(death ~ I(age-30) + I(tbsa-15)*inh_inj, data=data,
           family="binomial")

fit3 <- glm(death ~ I(age-30)*inh_inj + I(tbsa-15), data=data,
           family="binomial")

fit4 <- glm(death ~ I(age-30) + I(tbsa-15), data=data,
           family="binomial")

AIC(fit1)
AIC(fit2)
AIC(fit3)
AIC(fit4)
```

It seems like interaction terms can be kept in the model. Let's investigate the final model with Hosmer-Lemeshow test. Small p-value from Hosmer-Lemeshow test would indicate bad fit.

```{r}
hoslem.test(data$death, fitted(fit1))
```

Tests p-value is quite high now so we can't reject the model. Let's use the model to make predictions for test data. If probability to die is over 0,5 let's say that the prediction is that patient will die.

```{r}
#Take relevant predictors
test <- data.frame(death = test$death, age = test$age, tbsa = test$tbsa, inh_inj = test$inh_inj)

#Predictions
pred <- predict(fit1, newdata = test, se.fit=TRUE, type="response")

test <- cbind(test, probs = round(pred$fit,2))

test$preds <- as.numeric(test$probs > 0.5)

#Ratio of correct predictions
sum(test$death == test$preds)/nrow(test)
```

The final model could predict the outcome in the test data with approximately 91 % accuracy. This seems like a reasonably good result.

The models assumptions are that all observations are independent. That means in this case that there's no patients that got hospitalized from same burn accident. Relationships between predictors and log-odds should be linear. Predictors for the final model are age, total burnt surface area and their interactions with inhalation injury. Age is centered at 30 years and total burnt surface area at 15 % for easier interpretations. Model can be interpreted using odds and odds ratios (OR) and at the probability level.

```{r}
expb <- exp(coef(fit1))
expb
```

Main interest was effect of age. Let's see related odds and odds ratios.

```{r}
#OR between two patients with same tbsa and no inhalation injury with age
#difference of one year.
expb[2]

#OR between two patients with same tbsa and inhalation injury with age
#difference of one year.
expb[2]+expb[5]
```

This means that older patient is more likely to die in hospital and even more likely if inhalation injury is present. Intercept can be interpreted as the odds for a 30 year old patient with 15 % total burnt surface area and no inhalation injury to die in hospital.

```{r}
expb[1]
```

Let's examine the effect of total burnt surface area and inhalation injury with odds and odds ratios.

```{r}
#OR between patients of same age and no inhalation injury with one percentage
#point difference in total burnt surface area
expb[4]

#OR between patients of same age and inhalation injury with one percentage
#point difference in total burnt surface area
expb[4]+expb[6]

#OR between patients of same age and with same burnt surface area. Other has
#inhalation injury and other has not.
expb[3]
```

So higher total burnt surface area raises the odds to die in a hospital and even more if patient also has inhalation injury. Inhalation injury has major effect on odds to die in a hospital. However the 95 % confidence interval for the last OR is quite large but one can still say that there's a major effect.

```{r}
exp(confint(fit1))
```

Let's look at the interpretations at the probability level.

```{r}
b <- coef(fit1)
b

invlogit <- function(x){
  plogis(x)
}

ages <- seq(-20, 60, 10)

probs1 <- invlogit(b[1] + ages*b[2])
probs2 <- invlogit(b[1] + ages*(b[2]+b[5]) + b[3])

d1 <- data.frame(age = ages+30, no_inh_inj = probs1, inh_inj = probs2)

#Probabilities to die in a hospital for different ages with tbsa of 15%.
d1

tbsas <- seq(-5, 75, 10)

probs3 <- invlogit(b[1] + tbsas*b[4])
probs4 <- invlogit(b[1] + tbsas*(b[4]+b[6]) + b[3])

d2 <- data.frame(tbsa = tbsas+15, no_inh_inj = probs3, inh_inj = probs4)

#Probabilities for 30-year old patients with different tbsa to die in a 
#hospital.
d2
```

So to summarize the results it looks like age definitely has an effect to patients chances to survive the burn injuries. Older patients are more likely to die to their injuries than younger patients. Inhalation injury has also severe effect and even somewhat healthy young adults have fairly high probabilities to die if they have inhalation injury and no major burn injuries. Total burnt surface area also has an effect on chances to survive. The more severe the burns are the more likely the patient is to die in a hospital.
\newline













