---
title: "Demo 5"
author: "Jussi Kauppinen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 5.1

\begin{equation}
\begin{split}
L(\pi_1,...,\pi_k) &= \prod_{i=1}^k \binom{n_i}{y_i} \pi_i^{i_i} (1 - \pi_i)^{n_i-y_i}\\
l((\pi_1,...,\pi_k) &= \sum_{i=1}^k \left[ \log\binom{n_i}{y_i} + y_i \log \pi_i + (n_i - y_i) \log(1-\pi_i) \right]\\
&= \sum_{i=1}^k \left[ \log\binom{n_i}{y_i} + n_i \log (1-\pi_i) + y_i \log \left( \frac{\pi_i}{1-\pi_i}\right) \right]\\
&= \sum_{i=1}^k \left[ \log\binom{n_i}{y_i} - n_i \log (1+e^{x_i^T \beta}) + y_i x_i^T \beta \right]\\
\frac{\partial}{\partial \beta_j} l(\beta) &= \sum_{i=1}^k -n_i \frac{x_{ij}e^{x_i^T \beta}}{1+e^{x_i^T \beta}} + y_i x_i^T \beta\\
&= \sum_{i=1}^k x_{ij} \left( y_i - n_i \frac{e^{x_i^T \beta}}{1+e^{x_i^T \beta}} \right)\\
&= \sum_{i=1}^k x_{ij} ( y_i - n_i \pi_i )\\
\end{split}
\end{equation}



## Task 5.2

\begin{equation}
\begin{split}
D &= 2(l(y_1,...,y_k) - l(\hat y_1,...,\hat y_k))\\
&= 2\sum_{i=1}^k y_i \log y_i + (1 - y_i) \log (1 - y_i) - y_i \log \hat y_i - (1 - y_i) \log (1 - \hat y_i)\\
&= 2\sum_{i=1}^k y_i \log \left( \frac{y_i}{\hat y_i} \right) + (1 - y_i) \log \left( \frac{1-y_i}{1-\hat y_i} \right)\\
\end{split}
\end{equation}

## Task 5.3

```{r}
tul <- read.table("http://users.jyu.fi/~knordhau/GLM2/korvatul.txt", TRUE)

#tul$lapsia <- as.factor(tul$lapsia)

#fit <- glm(tul$tulehdus ~ tul$tupak + tul$lapsia, data=tul, family=binomial)

#summary(fit)

# response and design
y <- tul[,1]
X0 <- rep(1,length(y))
X1 <- tul[,2]
X2 <- (tul[,3] == 2)
X3 <- (tul[,3] == 3)
X <- cbind(X0, X1, X2, X3)

# Starting values for beta and eps
beta <- c(-0.5, 0.2, 0.7, 1.5) #Muuta näitä tarpeen mukaan
eps <- 0.1
while (eps > 1e-4) {

# eta, mu and score function
eta <- X %*% beta
mu <- exp(eta) / (1 + exp(eta))
S <- t(X) %*% (y - mu)

# W matrix
Wii <- mu/(1 + exp(eta))
W <- diag(c(Wii))
# Fisher information matrix
I <- t(X) %*% W %*% X
# Iteration step
betanew <- beta + solve(I) %*% S
# performance measure
eps <- t(betanew-beta) %*% (betanew-beta)
# updating the estimate
beta <- betanew
}
# return a list of the coefficients and their standard errors
list(beta = beta, se = diag(sqrt(solve(t(X) %*% W %*% X))))
```


## Task 5.4

```{r}
bw <- read.table("http://users.jyu.fi/~knordhau/GLM2/lapset.txt", header=TRUE)
head(bw)

bw$lapsia <- cut(bw$parit, c(-1,0,1,20), labels=c(0,1,2))
bw$kesk <- as.numeric(bw$syntpaino < 2500)

plot(bw$kesk ~ bw$tup)

#Mean-weights for each smoking type of the mother
aggregate(bw$syntpaino, list(bw$tup), data=bw, FUN=mean)

smoke0 <- bw[bw$tup == 0,]
smoke1 <- bw[bw$tup == 1,]
smoke2 <- bw[bw$tup == 2,]

#Frequencies of premature babies for each smoking group divided by
#number of observations
sum(smoke0$kesk)/nrow(smoke0)
sum(smoke1$kesk)/nrow(smoke1)
sum(smoke2$kesk)/nrow(smoke2)

#According to these statistics it would seem like smokers have a higher risk
#of having premature baby

#One could assume that social class could affect mothers smoking.

```


## Task 5.5

```{r}
bw$tup <- as.factor(bw$tup)

fit1 <- glm(bw$kesk ~ bw$tup * bw$lapsia, data = bw, family=binomial)
summary(fit1)

fit2 <- glm(bw$kesk ~ bw$tup + bw$lapsia, data = bw, family=binomial)
summary(fit2)

#Difference of deviances
anova(fit2, fit1)

#Test statistic is 2.5685 and difference in number of parameters is 2
1 - pchisq(2.5685,4)

#p-value is 0,63 indicates that smaller model is true

exp(coef(fit2))

#OR-between smoker types 1 and 0 is 1.40 when other factors are the same

#OR between smoker types 2 and 0 is 1.92 when other factors are the same

#OR between second child and first child is 0.47 when other factors are the same

#OR between later than second child and first child is 0.76 
#when other factors are the same

#odds for premature baby is thus higher the more the mother smokes
#when other factors are the same

#odds for child to be premature is lower for those with
#more one has previous pregnancies when other factors are the same

#confidence intervals
exp(confint(fit2))

```


## Task 5.6

## Task 5.7


