---
title: "Demo 4"
author: "Jussi Kauppinen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 4.1

a)  


\begin{equation}
\begin{split}
L(\mu, \sigma^2) &= \left(\frac{1}{\sqrt{2\pi\sigma^2}}\right)^n exp(-\frac{1}{2\sigma^2} \sum_{i=1}^n (y_i - \mu)^2)\\
l(\mu, \sigma^2) &= n \log \left( \frac{1}{\sqrt{2\pi\sigma^2}} \right) - \frac{1}{2\sigma^2} \sum_{i=1}^n (y_i - \mu)^2\\
S(\mu) &= - \frac{1}{2\sigma^2} \sum_{i=1}^n \frac{\partial}{\partial \mu} (y_i - \mu)^2\\
&= - \frac{1}{2\sigma^2} \sum_{i=1}^n -2(y_i - \mu)\\
&= \frac{1}{\sigma^2} \sum_{i=1}^n (y_i - \mu)\\
\end{split}
\end{equation}

b)  


\begin{equation}
\begin{split}
L(\lambda) &= e^{-\lambda n} \frac{\lambda^{\sum_{i=1}^n y_i}}{\prod_{i=1}^n y_i !}\\
l(\lambda) &= -\lambda n + \log \lambda \sum_{i=1}^n y_i - \sum_{i=1}^n \log y_i !\\
S(\lambda) &= \frac{1}{\lambda} \sum_{i=1}^n y_i - n\\
\end{split}
\end{equation}


## Task 4.2

```{r include=TRUE, eval=TRUE}
#a

# Score function for the normal distribution
# where the root should be at the origin
ScN <- function(y, mu, n) { n * (mean(y) - mu) }


plotScoresN <- function(n){
  y <- rnorm(n, 0, 1)
  mu <- seq(-2, 2, 0.1)
  plot(mu, ScN(y, mu, n), type = "l", xlab = "mu", ylab = "S(mu)")
  abline(0,0)
  for (i in 1:19) {
  y <- rnorm(n, 0, 1)
  lines(mu, ScN(y, mu, n))
  }
}

par(mfrow = c(2, 2))
plotScoresN(10)
plotScoresN(100)
plotScoresN(1000)
plotScoresN(10000)

#b

ScP <- function(y, lambda, n) { n * (mean(y)/lambda - 1) }

plotScoresP <- function(n){
  y <- rpois(n, 4)
  lambda <- seq(2, 6, 0.1)
  plot(lambda, ScP(y, lambda, n), type = "l", xlab = "lambda", ylab = "S(lambda)")
  abline(0,0)
  for (i in 1:19) {
  y <- rpois(n, 4)
  lines(lambda, ScP(y, lambda, n))
  }
}

par(mfrow = c(2, 2))
plotScoresP(10)
plotScoresP(100)
plotScoresP(1000)
plotScoresP(10000)

```
As the sample size gets larger the score functions values vary less. This can be explained by central limit theorem since both score functions here include the mean of the samples.

## Task 4.3

```{r}

densP <- function(n){
  scores <- numeric(100)
  for (i in 1:100){
    y <- rpois(n, 4)
    scores[i] <- ScP(y, 4, n)
  }
  
  plot(density(scores), main=paste("Sample size: ", n))
}

par(mfrow = c(2, 2))
densP(10)
densP(100)
densP(1000)
densP(10000)

```

We notice from the scale of the y-axis that the density plot gets flatter as sample size gets larger. One could also say that score functions values distribute more evenly.


